Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2022-09-02 paper/paper-infer.tex
save

@@

We then investigate previous estimates using the extended set of HapMap3 variants proposed here (Methods). As expected, compared to HapMap3, higher $h^2$ and lower $p$ estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}). 
A much larger $h^2$ estimate is obtained for lipoprotein(a) concentration [TODO: ADD NUMBERS], which is also reflected in a larger predictive performance.
[TODO: COMPUTE AVERAGE R2 BOOST FOR OTHER PHENO]
Interestingly, when using this extended set of variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
However, fitting with this extended set of variants is more computationally demanding [TODO RATIO]; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}).

@@

@@

@@

Here we introduce a few changes to LDpred2-auto, which makes it better at inferring these important parameters.
First, we extend LDpred2-auto with a third parameter $\alpha$ that controls the relationship between minor allele frequencies (or equivalently, standard deviations) of genotypes and expected effect sizes; the model becomes
\begin{equation}\label{eq:new_model}
\beta_j = S_j \gamma_j \sim \left\{
\begin{array}{ll}
\mathcal N \big( 0,~\sigma_\beta^2 \cdot (S_j^2)^{(\alpha + 1)} \big) & \mbox{with probability $p$,} \\
0 & \mbox{otherwise.}
\end{array}
\right.
\end{equation}
Therefore, it was earlier assumed that $\alpha = -1$ and $\sigma_\beta^2 = h^2 / (M p)$ in equation \eqref{eq:prev_model}. 
This new model in equation \eqref{eq:new_model} is similar to the model assumed by SBayesS, where $\alpha$ is called $S$ \cite{zeng2021widespread}. 
In SBayesS, they estimate $\alpha$ and $\sigma_\beta^2$ by maximizing the likelihood of the normal distribution (over the causal variants from the Gibbs sampler).
In the new LDpred2-auto, we first sample causal variants with replacement (bootstrap) before finding the maximum likelihood estimators, such that we add some proper sampling to these two parameters. 
This maximum likelihood estimation is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be within -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the previous estimate.
We now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more sampling in order to account for the reduced effective number of correlated variants.
As for $h^2$, we still estimate it by $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants. We constrain this estimate to be at least 0.001 to prevent the Gibbs sampler from being trapped in very small heritability estimates.

@@

Authors thank members of the NCRR/QGG StatGen group for helpful discussions.
Authors also thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
This research has been conducted using the UK Biobank Resource under Application Number 58024.

@@

>>>>

[OVERVIEW TABLE OF WHAT LDPRED2 CAN PROVIDE?]

@@

We then investigate previous estimates using the extended set of HapMap3 variants proposed here (Methods). As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [10.8-13.7]) and lower $p$ (decrease of 11.5\% [10.6-12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}). 
This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1-8.2]).
A particularly much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502-0.514] instead of 0.324 [0.320-0.329]), which is also reflected in a larger predictive performance ($r^2$ in test set of 0.516 [0.508-0.524] instead of 0.344 [0.335-0.353]).
Interestingly, when using this extended set of variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
However, fitting with this extended set of variants is more computationally demanding [TODO RATIO]; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}).

Finally, we investigate different transformations to apply to some continuous phenotypes used here.
Indeed, 49 of such phenotypes seems log-normally distributed or heavy-tailed (when visualizing their histogram); we therefore log-transform them.
However, we do investigate alternative transformations here to decide which one should be preferred and to check the impact on the inference from LDpred2-auto.
Note that we use the HapMap3+ set of variants here.
We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates for $p$ and $\alpha$ are highly consistent.
However, $h^2$ estimates and predictive performance $r^2$ (in the test set) are generally larger with the log-transformation, hinting that it probably makes sense to transform these phenotypes.
We then compare to using the rank-based inverse normal (RIN) transformation in Figure \ref{fig:RINT}; estimates for $p$ and $\alpha$ are also highly consistent.
Except for bilirubin and lipoprotein(a) concentration, generally higher $h^2$ estimates and predictive performance $r^2$ are obtained with the RIN-transformation than the log-transformation.

\subsection*{Local heritability and poligenicity}

In this section, we use the new HapMap3+ set of variants constructed here. 
We recall that we have defined 431 independent LD blocks for this set (Methods).
We compute a per-block (local) $h^2$ estimate, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks.

@@

[HM3\_PLUS AND MISSING HERITABILITY]

@@

[FURTHER RESTRICT VARIANTS THAT CAN ENTER THIS NEW SET?]

@@

Here we introduce a few changes to LDpred2-auto, which makes it better at inferring these important parameters.
First, we extend LDpred2-auto with a third parameter $\alpha$ that controls the relationship between minor allele frequencies (or equivalently, standard deviations) of genotypes and expected effect sizes; the model becomes
\begin{equation}\label{eq:new_model}
\beta_j = S_j \gamma_j \sim \left\{
\begin{array}{ll}
\mathcal N \big( 0,~\sigma_\beta^2 \cdot (S_j^2)^{(\alpha + 1)} \big) & \mbox{with probability $p$,} \\
0 & \mbox{otherwise.}
\end{array}
\right.
\end{equation}
Therefore, it was earlier assumed that $\alpha = -1$ and $\sigma_\beta^2 = h^2 / (M p)$ in equation \eqref{eq:prev_model}. 
This new model in equation \eqref{eq:new_model} is similar to the model assumed by SBayesS, where $\alpha$ is called $S$ \cite{zeng2021widespread}. 
In SBayesS, they estimate $\alpha$ and $\sigma_\beta^2$ by maximizing the likelihood of the normal distribution (over the causal variants from the Gibbs sampler).
In the new LDpred2-auto, we first sample causal variants with replacement (bootstrap) before computing the maximum likelihood estimators, such that we add some proper sampling to these two parameters. 
This maximum likelihood estimation is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration.
We now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more sampling by properly accounting for the reduced effective number of correlated variants.
As for $h^2$, we still estimate it by $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants and $\boldsymbol{\beta}$ is a vector of causal effect sizes (after scaling) from one iteration of the Gibbs sampler. 
We constrain this estimate to be at least 0.001 to prevent the Gibbs sampler from being trapped in very small heritability estimates.
Note that this $h^2$ estimate can be restricted to e.g.\ variants from a single LD block to get estimates of local heritability.

@@

Authors thank members of the NCRR/QGG StatGen group for helpful discussions.
Authors also thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
This research has been conducted using the UK Biobank Resource under Application Number 58024; authors thank all the participants of the UK Biobank for contributing to such useful data for Research.

@@

[TODO: TABLE EXPORTING RESULTS (ESTIMATES)]
